### 2023-08-29 sections/2-methods.tex
dataset reordering

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point in the dataset, 
which we define as: 
$$\text{log}_2 \left( \frac{|B_X(q, r_1)|}{|B_X(q, r_2)|} \right)$$
where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ 
centered at a point $q$ in the dataset $\textbf{X}$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold within the 8,575-dimensional embedding space. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold at a given point, for some length scale.
The notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold is known as the \emph{manifold hypothesis}.
[TODO: add example for GreenGenes and add lfd plots from CHESS]

@@

[TODO: Do set comprehension definitions for these and then prose refinement]
With these concepts defined, we can now pose the $k$- and $\rho$ nearest neighbors search problems.
Given a query $q$, along with a distance function $f$ defined on a dataset $ \textbf{X}$, $k$-nearest neighbors search aims to: 
for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$; that is, find the $k$ points $x \in \textbf{X}$ with the smallest $f(q, x)$.
We also have the $\rho$-nearest neighbors search problem: for a given radius $\rho$, find all points in $\textbf{X}$ that are a distance of $\rho$ 
or less from $q$; that is, all points $x \in \textbf{X}$ such that $f(q, x) \leq \rho$.

@@

[TODO: talk about singleltons or whatever and overloading of notation]
We sometimes also use this notation with points instead of clusters. For example, we sometimes use $\delta = f(x, q)$ to refer to the distance 
from a point $x$ to the query $q$. Since we can think of a point as having radius $r=0$, for points we have that $\delta_{max} = \delta = \delta_{min}$.

@@

[TODO: take out memoization and talk about reordering instead]
We start by building a divisive hierarchical clustering of the dataset with CLAM, using a 
similar recursive procedure as outlined in CHESS~\cite{ishaq2019clustered}, but with the following 
improvements: better selection of poles for partitioning and memoization of useful cluster properties, 
such as local fractal dimension (see Section 2.4.2: $k$-NN by Repeated $\rho$-NN).

@@

[TODO: use cluster C and |C| instead here; make sure you mention how radius is calculated; sampling, center, radius, poles; 
define geometric median]
For a cluster with $m$ points, we begin by computing a 
random sample of $\sqrt m$ points. We then compute the geometric median of this sample, which we call the 
cluster's center. We then compute the distance from the center to each point in the sample of $\sqrt m$ points. 
The point $l$ which is furthest from the center is designated the left pole, and the point $r$ which is furthest
from $l$ is designated the right pole. We then partition the cluster into a left child and right child, where the 
left child contains all points in the cluster which are closer to $l$ than to $r$, and the right child contains all 
points in the cluster which are closer to $r$ than to $l$. Without loss of generality, we assign to the left child 
those points which are equidistant from $l$ and $r$. Starting from a root-cluster containing the entire dataset, we 
repeat this procedure until each leaf contains only one datum or until some other user-specified stopping criterion 
is met.

@@

\subsubsection {Dataset Reordering}

[2 tiers of improvement]

CLAM-CAKES also reorders the dataset to improve the tree-building process 
used in CHESS. In CHESS, each cluster stored a vector of indices into the dataset, which 
were used to retrieve its points during search. Thus, for a dataset with cardinality $n$, 
at each depth in the tree, we would store a total of $n$ indices. Since we have approximately

@@

[TODO: delete this and talk about overlapping children improvement instead]
We conduct $\rho$-nearest neighbors search similarly as described in CHESS~\cite{ishaq2019clustered}, but 
with some implementation improvements including identification of candidate clusters which are leaves and 
conducting exhaustive search over those leaves.

>>>>

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point in the dataset, 
which we define as: 
\begin{equation} \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) } \label{1} \end{equation}
where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ 
centered at a point $q$ in the dataset $\textbf{X}$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold within the 8,575-dimensional embedding space. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold at a given point, for some length scale.
The notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold is known as the \emph{manifold hypothesis}.
[TODO: add example for GreenGenes and add lfd plots from CHESS]

@@

With these concepts defined, we can now pose the $k$- and $\rho$ nearest neighbors search problems.
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $\{x \in \textbf{X}: f(q, x) \leq f(q, x_k) \land x_k \text{ is the $k$th nearest neighbor of $q$ in $\textbf{X}$}\}$; that is,
for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
We also have the $\rho$-nearest neighbors search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$; that is, 
find all points in $\textbf{X}$ that at most a distance $\rho$ from $q$.

@@

We define \emph{singletons} as clusters which either contain a single point (i.e., have cardinality 1) or which contain 
many instances of the same point (i.e., have cardinality greater than 1 but contain only one unique point). Singletons clearly 
have zero radius, and so $\delta_{max} = \delta_{min} = \delta$ for these clusters. Hence, we sometimes overload the above 
notation to refer to distance from a query to a point. Similarly, we have that $\delta_{max} = \delta = \delta_{min}$ for 
individual points.

@@

We start by building a divisive hierarchical clustering of the dataset with CLAM, using a 
similar recursive procedure as outlined in CHESS~\cite{ishaq2019clustered}, but with the following 
improvements: better selection of poles for partitioning and depth-first dataset reordering 
(see Section ADD SECTION: Dataset Reordering).

@@

For a cluster $C$ with $|C|$ points, we begin by computing a 
random sample of $\sqrt{|C|}$ points, and computing pairwise distances 
between all points in the sample. Based on those pairwise distances, we compute the \emph{geometric median} of this sample; 
that is, the point which minimizes the sum of distances to all other points in the sample. This geometric median 
is $C$'s \emph{center}. The \emph{radius} of $C$ is the maximum distance from any point in $C$ to its center.
The point $l$ which is furthest from the center is designated the \emph{left pole}, and the point $r$ which is furthest
from $l$ is designated the \emph{right pole}. We then partition the cluster into a \emph{left child} and \emph{right child}, where the 
left child contains all points in the cluster which are closer to $l$ than to $r$, and the right child contains all 
points in the cluster which are closer to $r$ than to $l$. Without loss of generality, we assign to the left child 
those points which are equidistant from $l$ and $r$. Starting from a root-cluster containing the entire dataset, we 
repeat this procedure until each leaf contains only one datum or until some other user-specified stopping criterion 
is met.

@@

\subsubsection {Dataset Depth-First Reordering}

CLAM-CAKES also reorders the dataset to improve the tree-building process 
used in CHESS. In CHESS, each cluster stored a vector of indices of its points. 
We use these indices during search to retrieve a cluster's points from the dataset. 
Though this approach allowed us to retrieve the points in constant 
time, its memory cost was prohibitively high; since each cluster stores indices of all of its
points, for a dataset with cardinality $n$, we store a total of $n$ indices at each depth in the tree.
With $\log{}n$ levels in the tree, this approach had $\mathcal{O}(n\log{}n)$ memory cost.

A later improvement reduced the memory cost to $\mathcal{O}(n)$ by storing only the indices of the points in the leaves.
This approach, however, introduced an $\mathcal{O}(n)$ time cost every time we needed to find the indices for 
a non-leaf cluster, since this required a traversal of the subtree rooted at that cluster.

In this work, we introduce a new approach, wherein after building the cluster tree, we reorder the dataset 
so that points are stored in a depth-first order based on the tree. Then, within each cluster, we need only 
store the cluster's cardinality, and an \emph{offset} to access the cluster's points from the dataset.
For left leaf clusters, the offset is the index of the first point in the cluster. Right leaves have an offset 
equal to their sibling's offset plus the their sibling's cardinality. Parents have the same offset as their left child.
With $\mathcal{O}(n)$ memory cost and $\mathcal{O}(1)$ time cost for retrieving a points during search, 
dataset depth-first reordering offers significant improvements over the previous approaches.

@@

[TODO: delete this and talk about overlapping children improvement instead]
We conduct $\rho$-nearest neighbors search as described in CHESS~\cite{ishaq2019clustered}, but 
with some implementation improvements including identification of candidate clusters which are leaves and 
conducting exhaustive search over those leaves.
